---
title: "Lasso-Default Washington D.C"
author: "Hashan Fernando"
date: "2024-01-31"
output: 
  html_document: 
    theme: lumen
---

```{r, include=FALSE}
library(sf)
library(mgcv)
library(PerformanceAnalytics)
library(dplyr)
library(ggplot2)
library(GGally) 
library(gratia)
library(caret)
library(Metrics)
library(glmnet)
library(reshape2)
```

```{r}
# Importing the dataset
dcRawHepvu <- st_read("/Users/h6x/ORNL/git/opioid-risk-index/washington dc/data/processed data/SVI2020 WashingtonDC counties with death rate and count cdc/SVI2020_WashingtonDC_counties_with_death_rate_HepVu_count_cdc.shp")


# Getting the column names
column_names <- names(dcRawHepvu)
#print(column_names)

```

```{r}
print(column_names  )
```

```{r}
# creating new column to get the number od deaths from the rate
dcRawHepvu <- mutate(dcRawHepvu, DeathCount = round((NOD_Rate_2*E_TOTPOP)/100000))
```

```{r}
# print few number of coumns from the datset
data_comparison <- dcRawHepvu %>% 
  select(one_of(c("COUNTY","Deaths","od_deaths_","NOD_Rate_2","DeathCount")))

```

#### Let's study the response variable

```{r}
hist(dcRawHepvu$DeathCount, main = "Histogram of DeathCount", xlab = "DeathCount", col = "skyblue", border = "black",breaks=30)

```


```{r}
mean(dcRawHepvu$DeathCount)
var(dcRawHepvu$DeathCount)

```
This high variance relative to the mean may indicate overdispersion, which means that the variability in the data is greater than what would be expected under a simple Poisson model. Overdispersion is common in count data, and it might lead to issues if you try to fit a Poisson regression model directly.


```{r}

# Filtering the data set
desired_columns <- c("EP_POV150","EP_UNEMP","EP_HBURD","EP_NOHSDP","EP_UNINSUR","EP_AGE65","EP_AGE17","EP_DISABL","EP_SNGPNT","EP_LIMENG","EP_MINRTY","EP_MUNIT","EP_MOBILE","EP_CROWD","EP_NOVEH","EP_GROUPQ","DeathCount")

filtered_data <- dcRawHepvu %>%
  select(one_of(desired_columns))

filtered_data <- st_drop_geometry(filtered_data)
```

```{r}
filtered_data
```


```{r}
# Splitting the dataset into features (X) and target variable (y)

y <- filtered_data$DeathCount
x <- data.matrix(filtered_data[, !names(filtered_data) %in% "DeathCount"])

# Set a seed for reproducibility
set.seed(123)

# Create an index for splitting the data (70% train, 30% test)
splitIndex <- createDataPartition(y, p = 0.7, list = FALSE)

# Create training and testing sets
x_train <- x[splitIndex, ]
y_train <- y[splitIndex]
x_test <- x[-splitIndex, ]
y_test <- y[-splitIndex]

```

### Lasso Regression

To determine what value to use for lambda, we’ll perform k-fold cross-validation and identify the lambda value that produces the lowest test mean squared error (MSE).  cv.glmnet() automatically performs k-fold cross validation using k = 10 folds.


```{r}
### LASSO REGRESSION ###
m_lasso <- glmnet(x_train, y_train, alpha = 1)

# plot model coefficients vs. shrinkage parameter lambda
plot(m_lasso, xvar = "lambda", label = TRUE, xlab = expression(paste("Log(", lambda, ")")))

```
```{r}
#perform k-fold cross-validation to find optimal lambda value

cv_model <- cv.glmnet(x_train, y_train, alpha = 1)

#find optimal lambda value that minimizes test MSE
best_lambda <- cv_model$lambda.min
best_lambda
```

```{r}
#produce plot of test MSE by lambda value
plot(cv_model) 
```

```{r}
#find coefficients of best model
best_model <- glmnet(x_train, y_train, alpha = 1, lambda = best_lambda)
coef(best_model)
```

```{r}
#use fitted best model to make predictions
y_predicted <- predict(best_model, s = best_lambda, newx = x)
```

```{r}
#find SST and SSE
sst <- sum((y - mean(y))^2)
sse <- sum((y_predicted - y)^2)
```

```{r}
#find R-Squared
rsq <- 1 - sse/sst
rsq
```


```{r}

# prediction for the training set
m_lasso_cv_pred_train <- predict(best_model, x_train, s = "lambda.min")

# calculate RMSE on training set
print(paste("RMSE on training set:",
            rmse(m_lasso_cv_pred_train,
                  y_train)
)
)

# prediction for the test set
m_lasso_cv_pred_test <- predict(best_model, x_test, s = "lambda.min")

# calculate RMSE for the test data set
print(paste("RMSE on test set:",
            rmse(m_lasso_cv_pred_test,
                  y_test)
)
)

```



### Session Info:

```{r}
sessionInfo()
```
