---
title: "Lasso-Poisson TN"
author: "Hashan Fernando"
date: "2024-01-31"
output: 
  html_document: 
    theme: lumen
---

```{r, include=FALSE}
library(sf)
library(mgcv)
library(PerformanceAnalytics)
library(dplyr)
library(ggplot2)
library(GGally) 
library(gratia)
library(caret)
library(Metrics)
library(glmnet)
library(reshape2)
```

```{r}
# Importing the dataset
tnRawHepvu <- st_read("/Users/h6x/ORNL/git/opioid-risk-index/tennessee/data/processed data/SVI2020 TN counties with death rate HepVu/SVI2020_TN_counties_with_death_rate_HepVu.shp")


# Getting the column names
column_names <- names(tnRawHepvu)
#print(column_names)

```

```{r}
print(column_names  )
```

```{r}
# creating new column to get the number od deaths from the rate
tnRawHepvu <- mutate(tnRawHepvu, DeathCount = round((NOD_Rate_2*E_TOTPOP)/100000))
```


#### Let's study the response variable

```{r}
hist(tnRawHepvu$DeathCount, main = "Histogram of DeathCount", xlab = "DeathCount", col = "skyblue", border = "black", breaks =30)

```


```{r}
mean(tnRawHepvu$DeathCount)
var(tnRawHepvu$DeathCount)

```
This high variance relative to the mean may indicate overdispersion, which means that the variability in the data is greater than what would be expected under a simple Poisson model. Overdispersion is common in count data, and it might lead to issues if you try to fit a Poisson regression model directly.


```{r}

# Filtering the data set
desired_columns <- c("EP_POV150","EP_UNEMP","EP_HBURD","EP_NOHSDP","EP_UNINSUR","EP_AGE65","EP_AGE17","EP_DISABL","EP_SNGPNT","EP_LIMENG","EP_MINRTY","EP_MUNIT","EP_MOBILE","EP_CROWD","EP_NOVEH","EP_GROUPQ","DeathCount")

filtered_data <- tnRawHepvu %>%
  select(one_of(desired_columns))

filtered_data <- st_drop_geometry(filtered_data)
```

```{r}
filtered_data
```

```{r}
# Splitting the dataset into features (X) and target variable (y)
X <- filtered_data[, !names(filtered_data) %in% "DeathCount"]  # Features
y <- filtered_data$DeathCount  # Target variable

# Set seed for reproducibility
set.seed(123)

# Create an index for splitting the data
index <- createDataPartition(filtered_data$DeathCount, p = 0.8, list = FALSE)


# Create training sets for features and target
X_train <- X[index, ]
y_train <- y[index]

X_train_matrix <- as.matrix(X_train)


# Create test sets for features and target
X_test <- X[-index, ]
y_test <- y[-index]


X_test_matrix <- as.matrix(X_test)
```



### Lasso Regression

To determine what value to use for lambda, we’ll perform k-fold cross-validation and identify the lambda value that produces the lowest test mean squared error (MSE).  cv.glmnet() automatically performs k-fold cross validation using k = 10 folds.


```{r}
### LASSO REGRESSION ###

m_lasso_cv <- cv.glmnet(X_train_matrix, y_train, alpha = 1,family = "poisson")

plot(m_lasso_cv)

m_lasso_cv$lambda.min

m_lasso_cv$lambda.1se


```

```{r}
#my
fit <- glmnet(X_train, y_train, alpha = 1,lambda = m_lasso_cv$lambda.min, family = "poisson")

fit$beta[,1]


# prediction for the training set
m_lasso_cv_pred_train <- predict(m_lasso_cv, X_train_matrix, s = "lambda.min")

# calculate RMSE on training set
print(paste("RMSE on training set:",
            rmse(m_lasso_cv_pred_train,
                  y_train)
)
)

# prediction for the test set
m_lasso_cv_pred_test <- predict(m_lasso_cv, X_test_matrix, s = "lambda.min")

# calculate RMSE for the test data set
print(paste("RMSE on test set:",
            rmse(m_lasso_cv_pred_test,
                  y_test)
)
)


```

### Session Info:

```{r}
sessionInfo()
```
